---
title: "Birth Defects in Minnesota Data Report"
---


How does food policy, education, and socioeconomic status shape birth defect outcomes across populations?

## Background & Motivation
Birth defects affect thousands of families in Minnesota and these conditions vary significantly in prevalence across racial, economic, and educational groups, from cardiovascular malformations to neural tube defects. Two questions drove forth our analysis for this project:

- Question 1: What maternal characteristics increase the risk for different types of birth defects?
- Question 2: How did the folic acid fortification policy affect the prevalence of Neural Tube Defects (NTDs)?

Neural tube defects (NTDs) are a serious condition that affect the brain, spine, or spinal cord and typiclly form within the first month of pregnancy, often before a mother even knows she is pregnant.  

Folate is a vitamin B9 nutrient that is important for DNA synthesis, cell division, and cell formation. Natural folate is found in leafy greens, legumes, nuts, and fortified grains, while synthetic folic acid is better absobred and typically taken as supplements. More than half of NTDs are preventable with adequate folate and folic acid intake.

The CDC recommends that all women from ages 15 to 45 should consume 400 mcg of folic acid daily, even if they are not planning on a pregnancy. 


## Folic Acid Fortification Policy
The path from scientific discovery to food policy was slow and contested — with real consequences for birth defect rates.

During the time of 1983–199, a landmark study found about a 71% reduction in NTDs when folic acid intake was increased. Thus, the CDC issued a formal recommendation, but supplements were only available by prescription which limited access for families of lower-income.

Then in 1993, the FDA began to consider fortification and added vitamins directly to grain products, although they recognized that supplements by themselves would not be able to reach populations who were most at risk, as well as the fact NTDs would form before many women knew they are pregnant.

Within the few years of 1996–1997, the FDA published its ruling on folic acid fortification and NTD rates begin to slowly decline slightly.

Then finally in 1998 and onwards, the folic acid fortification for grain products became mandatory and NTD rates continued to decline across all racial groups, although racial gaps did still persist.


## Supplementation: Education & Poverty Gaps
Despite the fortification policy, the daily supplementation rates varied drastically by education and poverty levels among Minnesota mothers (2016-2022). 

```{r}
library(tidytext)

folic_master %>%
  filter(
    Subtopic == "Folic Acid Use",
    Variable %in% c("Education Level", "Poverty Level")
  ) %>%
  group_by(Variable, Category) %>%
  summarize(pct = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(Category = reorder_within(Category, pct, Variable)) %>%
  ggplot(aes(x = pct, y = Category, color = Variable)) +
  geom_segment(aes(x = 0, xend = pct, yend = Category), linewidth = 0.8, color = "grey80") +
  geom_point(size = 6) +
  geom_text(aes(label = paste0(round(pct, 1), "%")), nudge_x = 5, size = 3,
            fontface = "bold", show.legend = FALSE) +
  scale_y_reordered() +
  scale_color_manual(values = c("Education Level" = "#185FA5", "Poverty Level" = "#85B7EB")) +
  scale_x_continuous(limits = c(0, 72), breaks = seq(0, 60, 10),
                     labels = scales::label_percent(scale = 1)) +
  facet_grid(Variable ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Folic Acid Supplementation by Socioeconomic Factors, 2016-2022",
    x = "Percent Taking Daily Folic Acid", y = NULL, color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 11, hjust = 0),
    strip.background = element_rect(fill = "grey95", color = NA),
    plot.title.position = "plot",  
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.text.y = element_text(size = 10, color = "grey20"),
    axis.text.x = element_text(color = "grey50"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey92", linewidth = 0.4),
    panel.spacing = unit(1, "lines")
  )
```

The gap between the lowest (21.3%) and the highest (53.2%) education groups is around 32.6%. This directional pattern is the same for poverty levels as mothers with low poverty levels supplement daily around 49.6% while those in more moderate-to-high poverty level groups hover around 24% of daily folic acid intake.
Both of the dimensions on education

point in the same direction, suggesting a compounding disadvantage. Mothers who are both lower-income and less educated face the greatest barriers to consistent folic acid intake — and therefore the highest exposure to preventable NTD risk.





## Birth Defect Prevalence by Race/Ethnicity
When looking at the birth defect outcomes by maternal race/ethnicity in Minnesota (2016–2020), the disparities reflected smth smth
```{r}
maternal_stacked <- maternal_master %>%
  filter(
    Variable == "Maternal race/ethnicity",
    !Category %in% c("All races/ethnicities combined", "Other/Unknown")
  ) %>%
  group_by(Category, Note) %>%
  summarize(avg_prevalence = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(Note = str_remove(Note, " \\(.*\\)"), Category = str_remove(Category, ",? Non-Hispanic| Non-Hispanic|/Non-Hispanic")) %>%
  group_by(Category) %>%
  mutate(total = sum(avg_prevalence)) %>% 
  ungroup()

ggplot(maternal_stacked, aes(x = reorder(Category, total),
                              y = avg_prevalence, 
                              fill = Note)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d(option = "viridis") +
  labs(
    title = "Birth Defect Prevalence by Race/Ethnicity & Defect Type",
    subtitle = "Per 10,000 live births, 2016–2020",
    x = NULL,
    y = "Total prevalence per 10,000 live births",
    fill = "Defect type"
  ) +
  theme_minimal() +
  theme(legend.position = "right",
        legend.text = element_text(size = 8),
        plot.title = element_text(face = "bold", size = 13),
        legend.key.size = unit(0.4, "cm")
        )
```

American Indian or Alaska Native and Black/non-Hispanic mothers have the highest burden in overall birth defect. These groups are those that also face disproportionate socioeconomic barriers, which is consistent with the supplementation gaps we observed for those of lower education and higher poverty levels.


```{r}
rates_clean <- rates %>%
  select(Year, Value, Category, Note) 

  ggplot(rates_clean, aes(x = Year, y = Value, shape = Category)) +
    annotate("rect", xmin = 1995, xmax = 1997, ymin = 2, ymax = 12, alpha = 0.5, fill = "purple") +
    annotate("rect", xmin = 1997, xmax = 2011, ymin = 2, ymax = 12, alpha = 0.5, fill = "yellow") +
    geom_point() +
    geom_smooth(se = FALSE) +
    theme_classic()
```
For neural tube defects specifically, Hispanic populations are seen to experience the highest NTD prevalence in Minnesota, even despite the racial gap narrowing after the 1998 fortification. Meanwhile, black/non-Hispanic mothers show the lowest rates of NTDs.


## Key Findings & Takeaways





### Data Notes & Limitations
The NTD prevalence data is only available for Hispanic, White/non-Hispanic, and Black/non-Hispanic groups, so the NTD and overall birth defect datasets cannot be combined directly as they may use different scaling methods and time periods.

